---
title: "Data Mining course 2020/2021"
author: "Andrzej Janusz"
output:
  html_notebook:
    df_print: paged
    fig_height: 10
    fig_width: 10
    rows.print: 10
  html_document:
    df_print: paged
subtitle: Lab3 - mining association rules
email: a.janusz@mimuw.edu.pl
---

### The plan:

0. Importing transaction data.
1. Implementations of frequent set and association rule mining algorithms in R - __arules__ package.
2. Quality measures for assciation rules.
3. Vizualization and exploration of rule sets in R.  

```{r setup, results=FALSE}
options(width = 120)
library(data.table)

# installation of the required packages (uncomment the line below):
# install.packages(c("arules", "arulesViz"))
library(arules)
library(arulesViz)
```

### Reading PoS data

There are multiple ways you can read data in the transaction format. The easiest is to read transactions directly from a text file using the _read.transactions_ function, implemented in the _arules_ package. It can read transactions in both of the formats discussed during our previous class:  

 - the _"basket"_ format - each line of a file contains all item ids belonging to the corresponding transaction,
 - the _"single"_ format - a 2d table where each transaction can be divided into multiple lines, one item per line.
 
 You may also transform any transaction data to the required format using _data.table_ library.


```{r transaction_data_reading}
# let's get back to the exemplary PoS data from previous class
posData = data.table::fread("sample_pos_data.csv", header = TRUE)
head(posData)

# we can read this data directly as transactions:
transactions = arules::read.transactions("sample_pos_data.csv", header = TRUE, format = "single", 
                                         sep = ",", cols = c("transactionID", "productID"))
inspect(head(transactions))

# now, we may pring a summary of the data:
summary(transactions)

# you may also transform transactions into a binary matrix...
transM = as(transactions, "matrix")
is(transM)
dim(transM)
head(colnames(transM))

# and a binary matrix back into transactions:
new_trans = as(transM, "transactions")
is(new_trans)
```

### Finding frequent itemsets

The _arules_ package allows to efficiently find all frequent itemsets using two different algorithms. The first one is the _apriori_ algorithm which we discussed during our previous meeting. The second one is called _ECLAT_ (Equivalence Class Transformation). You may read more about it in the paper:  

Zaki, M. J. (2000). _"Scalable algorithms for association mining"_. IEEE Transactions on Knowledge and Data Engineering. 12 (3): 372–390.  

While the _apriori_ algorithm searches for frequent itemsets in a breadth-first fashion, ECLAT is an example of a depth-first approach, thus it can be more efficient for some data sets.

```{r frequent_itemsets}
# let's use the same data set as the last time
data(Epub)
summary(Epub)

# generating frequent itemsets using the apriori algorithm
# (this is an efficient implementation which uses prefix trees to speed up support counting):
minSupport = 0.001         # this is a quite low value but it is only used for a demonstration
frequentItemSets = apriori(Epub, 
                           parameter = list(supp = minSupport, 
                                            minlen = 2, 
                                            target = "frequent itemsets"),
                           appearance = NULL, control = list(verbose = FALSE))
summary(frequentItemSets)
frequentItemSets = sort(frequentItemSets, decreasing = TRUE, by = "support")

# 'inspect' is the method for displaying itemsets and rules:
inspect(frequentItemSets[which(size(frequentItemSets) == 3)])

# we may generate the same itemsets using the ECLAT algorithm:
frequentItemSetsEclat = eclat(Epub, parameter = list(supp = minSupport, 
                                                     minlen = 2, 
                                                     target = "frequent itemsets"),
                              control = list(verbose = FALSE))
summary(frequentItemSetsEclat)
frequentItemSetsEclat = sort(frequentItemSetsEclat, decreasing = TRUE, by = "support")

# let's check the itemset of size 3:
inspect(frequentItemSetsEclat[which(size(frequentItemSetsEclat) == 3)])
```
### Association rules

To create association rules from frequent itemsets, we need to set a value of additional quality criteria, i.e., __confidence__. If we define the support of an itemset $X$ in a set of transactions $T$ as $$supp(X) = \frac{\|\{t \in T: X \subseteq t\}\|}{\|T\|},$$ then the confidence of an association rule $X \Rightarrow Y$ is $$conf(X \Rightarrow Y) = \frac{supp(X \cup Y)}{supp(X)}.$$

In practice, we usually do not expect that the confidence of discovered association rules is going to be very high. Please note that even $conf = 0.1$ can be meaningful when the support of $X$ is relatively high, but the support of $Y$ is low. In such a case, we may discover that the presence of $X$ dramatically increases the probability of the occurrence of $Y$ - which can be valuable information (think about the problem of up-sell).  

For this reason, it is convenient to use other association rule quality metric. One that is particularily popular is called __lift__: $$lift(X \Rightarrow Y) = \frac{supp(X \cup Y)}{supp(X) \times supp(Y)} = \frac{conf(X \Rightarrow Y)}{supp(Y)}.$$

```{r association_rules}
# we may generate association rules from frequent itemsets:
eclatRules = ruleInduction(frequentItemSetsEclat, Epub, confidence = 0.6, 
                           control = list(method = "apriori", verbose = FALSE))

inspect(eclatRules)

# we may also compute the rules directly using the apriori function - let's compute some more rules
ruleSet = apriori(Epub, parameter = list(supp = minSupport/2, 
                                         conf = 0.2,
                                         minlen = 2,
                                         target = "rules"),
                  appearance = NULL, 
                  control = list(sort = -1, verbose = FALSE))

summary(ruleSet)

# we may sort the rules by any metric
ruleSet = sort(ruleSet, decreasing = TRUE, by = "lift")
inspect(ruleSet[1:5])

# to get a data.frame with quality metrics we can use the 'quality' method:
inspect(ruleSet[which.max(quality(ruleSet)[["support"]])])

head(quality(ruleSet))

# we may compute some more metrics if we want to:
ruleSet = apriori(Epub, parameter = list(supp = minSupport/2, 
                                         conf = 0.2, 
                                         minlen = 2, 
                                         arem = "chi2",
                                         target = "rules", 
                                         ext = TRUE,
                                         minval = 0,
                                         aval = TRUE),
                  appearance = NULL, 
                  control = list(sort = -1, verbose = FALSE))
summary(ruleSet)

# we can also compute only specific type of rules, e.g., those which have item "doc_71" in the right-hand side part
ruleSet_doc_71 = apriori(Epub, parameter = list(supp = minSupport/2, 
                                                conf = 0.2,
                                                minlen = 2,
                                                target = "rules"),
                         appearance = list(rhs = "doc_71"), 
                         control = list(sort = -1, verbose = FALSE))
inspect(ruleSet_doc_71)
```

It might be useful to know how to effectively explore an association rule set. 

```{r exploration, fig.hold='hold', out.width="100%"}
rules = ruleSet

# we may want to filter out some rules:
maxRules = 100

ruleSet = sort(ruleSet, by = "lift")
ruleSet = ruleSet[1:maxRules]
summary(ruleSet)

# we may apply some additional filters
ruleSet = subset(ruleSet, subset = ((confidence > 0.9 & lift > 200) | size(ruleSet) > 4))
summary(ruleSet)

# We may also explore the rule set interactively (it will not work in the notebook - try it yourself in R session):
ruleSet = rules
plot(ruleSet[1:1000], engine='htmlwidget', shading = "chi2")
```
```{r exploration_plotting, fig.hold='hold', out.width="100%", fig.width=10}
# some nice plots
plot(ruleSet, data = Epub, method = "grouped", engine='default', shading = "lift")

plot(ruleSet[1:100], data = Epub, method="graph", engine="htmlwidget", shading = "lift")

plot(ruleSet[1:100], data = Epub, method="matrix", engine="htmlwidget", shading = c("lift", "confidence"))
```

### Exercise:

A link to the cosmetics data set:  
https://drive.google.com/file/d/1dbpYjIN3ArP2NZgviGWNiTIDj3KoitCf/view?usp=sharing

```{r real_life_data}
# now, let's investigate some real-life PoS data
cosmeticsData = data.table::fread('cosmeticsData.csv', sep = ',', header = TRUE) # the data set is available in the Drive
str(cosmeticsData)

# some of the data, including product ids, is missing...
cosmeticsData[, any(is.na(productID))]
data.table::fwrite(cosmeticsData[!is.na(productID)], file = 'cosmeticsData_cleaned.csv', 
                   col.names = TRUE, sep = ',', na = 'NA')

# we may also read the data as transactions:
cosmeticsTransactions = arules::read.transactions('cosmeticsData_cleaned.csv', 
                                                  sep = ',', format = 'single', 
                                                  cols = c('transactionID', 'productID'),
                                                  rm.duplicates = TRUE, header = TRUE)
inspect(head(cosmeticsTransactions))
```


**The task:**  
Please investigate this data set and find interesting association rules. Play with data visualization.    
Can you discover any interesting patterns?    
For instance, please try to add to each transaction the ID of a sales person (if it is available). Can you find any interesting associations that involve the sales personnel?    
Can you discover any relation between sales and time?    

\
\
\
  
  
  